---
title: "Descriptive Report"
subtitle: "`r params$comparison`"
author: "Carlos S. Reyna-Blanco"
license: "CC-BY 4.0"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
    theme: united
    highlight: tango
    toc: true           # Enable table of contents
    toc_float: true     # Make the table of contents float on the side
    number_sections: true  # Optional: Number your sections
    toc_depth: 4  
params:
  metadata:      NULL    # will be a data.frame of samples × columns
  exist:         NULL    # will be a data.frame of peptides × samples
  comparison:    NULL    # the name of this comparison
  library_meta:  NULL    # will be a data.frame of the library metadata
  custom_colors: NULL    # Colors to use for each group
  timepoints:    NULL    # Only passed if data is longitudinal and a ind_timepoint.csv is given
  out_tables:    NULL    # where to save the tables
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo  = FALSE, warning = FALSE, message = FALSE
                      ,fig.width  = 5, fig.height = 4) # default figure dimensions (in inches)
# if  prefer pixel dimensions, you can also set dpi and out.width:
# dpi       = 96, out.width = "800px"

# Load custom utility functions
source(file.path("utils", "report_utils.R"))

# read in the two per‐comparison files
metadata  <- params$metadata
exist <- params$exist
comparison <- params$comparison
lib_metadata_df <- params$library_meta
custom_colors <- params$custom_colors
out_dir <- params$out_tables

# metadata      = meta_i
# exist         = exist_i
# comparison    = name
# lib_metadata_df  = lib_metadata_df
# custom_colors = custom_colors
          
actual_subgroup_columns <- setdiff(names(SUBGROUPS_TO_NAME), "all")
subgroup_lib_df <- lib_metadata_df %>%
  tibble::rownames_to_column(var = "Peptide") %>%
  mutate(
    across(
      all_of(actual_subgroup_columns),
      ~ case_when(
        is.na(.)                       ~ FALSE,
        . %in% c(1, "1", TRUE, "True") ~ TRUE,
        TRUE                           ~ FALSE
      )
    ),
    all = TRUE
  ) %>%
  select(Peptide, all_of(SUBGROUPS_TO_INCLUDE))
```


```{r}
# include couns and diversity
shannon_diversity <- vegan::diversity(t(exist), index = "shannon", base = 2)
samples_stats <- data.frame(SampleName = names(shannon_diversity), enriched_peptides = colSums(exist), Shannon_diversity = shannon_diversity)
metadata <- metadata %>% 
  merge(samples_stats, by = "SampleName") #, all.x = TRUE)

# find Sex and Age column names, case‑insensitive
lc <- tolower(names(metadata))
# find age and sex/gender columns
age_idx <- which(lc == "age")
sex_idx <- which(lc %in% c("sex", "gender"))

if (length(age_idx) == 1 && length(sex_idx) == 1) {
  age_col <- names(metadata)[age_idx]
  sex_col <- names(metadata)[sex_idx]
  
  metadata <- metadata %>%
    mutate(
      Age_group = cut(
        .data[[age_col]],
        breaks = seq(0, 120, by = 5),
        right  = FALSE
      ),
      
      Sex_ctg = {
        # coerce to character and lowercase once
        sex_raw <- as.character(.data[[sex_col]])
        sex_lc  <- str_to_lower(sex_raw)
        
        # map to “Female” or “Male”
        sex_norm <- case_when(
          sex_lc %in% c("female", "f", "0") ~ "Female",
          sex_lc %in% c("male",   "m", "1") ~ "Male",
          TRUE                              ~ NA_character_
        )
        
        # factor with the desired order
        factor(sex_norm, levels = c("Female", "Male"))
      }
    )
  
} else {
  warning("Could not find exactly one Age and one Sex/Gender column; skipping Sex/Age grouping.")
}

# set group and merge features with group test labels
group_cols <- c("group_test")
features_target <- as.data.frame(t(exist)) %>%
  tibble::rownames_to_column("SampleName")%>%
  dplyr::left_join(metadata %>% select(any_of(c("SampleName", group_cols))), by = "SampleName")
```


```{r, results='asis'}
required <- c("Age_group", "Sex_ctg")
show_sexAge <- all(required %in% colnames(metadata))
```


`r if (show_sexAge) "\n# Sex/Age Distribution"` 


```{r, fig.height=5, fig.width=8}
if (show_sexAge) {
  sex_colors <- c("Female" = "#66C2A5", "Male" = "#FC8D62")
  p <- plot_sex_age_distribution(data = metadata, 
                                 group_col = group_cols, 
                                 age_col = "Age_group", 
                                 sex_col = "Sex_ctg", 
                                 custom_colors = sex_colors)
  print(p)
} 
```


# Antibody Epitope Repertoire of Samples

```{r, fig.height=5, fig.width=7}
for (group_col in group_cols){
  p <- plot_enrichment_counts(
    features_target      = features_target,
    group_col            = group_col,
    group_cols           = group_cols,
    prevalence_threshold = 0,
    custom_colors        = custom_colors,
    binwidth             = 0.5
  )
  print(p)
}
```


# Enriched Peptides Distribution

```{r}
# Enriched peptides ----
values_col <- "enriched_peptides"
for (group_col in group_cols){
  pairwise_comparisons <- combn(levels(factor(metadata[[group_col]])), 2, simplify = FALSE)
  p <- plot_groups_boxplots(data = metadata, 
                            group_col = group_col, 
                            values_col = values_col,
                            custom_colors = custom_colors, 
                            pairwise_comparisons = pairwise_comparisons,
                            label_axis = c("Group Test", "Counts of Enriched Peptides"))
  print(p)
}
```


# Shanon Diversity

```{r}
values_col <- "Shannon_diversity"
for (group_col in group_cols){
  pairwise_comparisons <- combn(levels(factor(metadata[[group_col]])), 2, simplify = FALSE)
  p <- plot_groups_boxplots(data = metadata, 
                       group_col = group_col, 
                       values_col = values_col,
                       custom_colors = custom_colors, 
                       pairwise_comparisons = pairwise_comparisons,
                       label_axis = c("Group Test", "Shannon Diversity"))
  print(p)
}
```


# Group Peptide Comparisons {.tabset}

```{r}
num_samples_per_group <- list()
for (group_col in group_cols) {
  num_samples_per_group[[group_col]] <- metadata %>%
    filter(!is.na(!!sym(group_col))) %>%  # Filter out NA values in the group_test column
    group_by(!!sym(group_col)) %>%
    summarise(n = n(), .groups = "drop") %>%
    tibble::deframe()  # Convert the summarized result to a named vector (group -> count)
}

percentage_group_test_list <- list()
for (group_col in group_cols){
  percentage_group_test_list[[group_col]] <-  get_count_percentage_df(
    features_target      = features_target,
    group_col            = group_col,
    group_cols           = group_cols,
    lib_metadata_df      = lib_metadata_df,
    prevalence_threshold = 0)
}

prevalence_threshold <- 0
comparison_group_tests <- automate_group_test_analysis(percentage_group_test_list, num_samples_per_group, prevalence_threshold)
```


## Scatterplot

```{r, results='asis'}
#group <- names(comparison_group_tests)[1]
all_plots <- lapply(comparison_group_tests[[group_col]], function(x) {
  # Adjusting the plot margins to avoid hover text cropping
  x$plot <- htmltools::div(style = "margin-bottom: 150px;", x$plot)
})
htmltools::tagList(all_plots)
```
<div style="height: 50px;"></div>


## Volcano plot

```{r, results='asis'}
volcano_plots <- list()
for (group_col in group_cols) {
  comparisons <- names(comparison_group_tests[[group_col]])
  for (comp_name in comparisons) {
    #comp_name <- comparisons[1]
    groups <- strsplit(comp_name, "_vs_")[[1]]
    group1 <- groups[1]
    group2 <- groups[2]
    comparison_df <- comparison_group_tests[[group_col]][[comp_name]]$comparison_df
    p <- make_interactive_volcano(comparison_df = comparison_df, group1 = group1, group2 = group2)
    volcano_plots[[paste0(group1, "_vs_", group2)]] <- list(plot = p %>% htmltools::div(style = "margin-bottom: 150px;"))
    # ggsave(paste0(group_col, "_", comp_name, "_volcano_plot.png"), p)
  }
  print(htmltools::tagList(volcano_plots))
}
```



## Table

```{r, results='asis'}
DT::datatable(
  head(comparison_group_tests[[group_cols[1]]][[comparison]]$comparison_df, 300),
  options = list(
    pageLength = 5,
    scrollX = TRUE,  # enable horizontal scrolling
    width = "100%"
  ),
  caption = "Preview of the top 300 Peptides ordered by delta ratio and pvals"
)
```
<div style="height: 50px;"></div>



# Multidimensional Scaling (MDS)

```{r, results='asis'}
p <- plot_mds(features_target, group_cols[1], 
              group_cols = group_cols,
                     custom_colors,
                     method = "jaccard",
                     permutations = 999) 
# Return plotly
ggplotly(p, tooltip = "text") %>%
  layout(margin = list(l = 80, r = 80, b = 80, t = 80, pad = -2),
         height = 550,  # Define the plot height in pixels
         width = 600,
         hoverlabel = list(font = list(size = 10)))
```
<div style="height: 150px;"></div>


```{r, results='asis'}
show_timepoints <- !is.null(params$timepoints)
if (show_timepoints) {
  cat('\n\n<div style="height: 60px;"></div>\n\n')  # adds vertical space
  cat('\n\n<div style="height: 60px;"></div>\n\n')  # adds vertical space
}

```


`r if (show_timepoints) "\n\n# Longitudinal Similarity {.tabset}"`  
`r if (show_timepoints) "## Data in at least one timepoint"`  


```{r, fig.height=9, fig.width=9}
if (show_timepoints) {
timepoints_df <- params$timepoints
g1 <- levels(factor(metadata[[group_col]]))[1]
g2 <- levels(factor(metadata[[group_col]]))[2]
  
plot_correlation(
  phiseq_df = exist,           # assuming your function uses this as the data matrix
  metadata = timepoints_df,
  ind_id_col = "ind_id",
  samples_t1_col = g1,
  samples_t2_col = g2,
  label_x = g1,
  label_y = g2,
  method = "pearson", 
  require_both_timepoints = FALSE
  )
}
```


`r if (show_timepoints) "## Data in both timepoints"`  


```{r, fig.height=9, fig.width=9}
if (show_timepoints){
timepoints_df <- params$timepoints
g1 <- levels(factor(metadata[[group_col]]))[1]
g2 <- levels(factor(metadata[[group_col]]))[2]
  
plot_correlation(
  phiseq_df = exist,           # assuming your function uses this as the data matrix
  metadata = timepoints_df,
  ind_id_col = "ind_id",
  samples_t1_col = g1,
  samples_t2_col = g2,
  label_x = g1,
  label_y = g2,
  method = "pearson", 
  require_both_timepoints = TRUE
  )
}
```


`r if (show_timepoints) "\n\n#  Random and Matched Pairs of Samples Similarity Distribution {.tabset}"`  
`r if (show_timepoints) "## Data in at least one timepoint"`


```{r, fig.height=6, fig.width=6}
if (show_timepoints){
timepoints_df <- params$timepoints
g1 <- levels(factor(metadata[[group_col]]))[1]
g2 <- levels(factor(metadata[[group_col]]))[2]

    
plot_correlation_distribution(
  phiseq_df = exist,           # assuming your function uses this as the data matrix
  metadata = timepoints_df,
  ind_id_col = "ind_id",
  samples_t1_col = g1,
  samples_t2_col = g2,
  label_x = g1,
  label_y = g2,
  method = "pearson", 
  require_both_timepoints = FALSE
  )
}
```


`r if (show_timepoints) "## Data in both timepoints"`  


```{r, fig.height=6, fig.width=6}
if (show_timepoints){
timepoints_df <- params$timepoints
g1 <- levels(factor(metadata[[group_col]]))[1]
g2 <- levels(factor(metadata[[group_col]]))[2]
 
plot_correlation_distribution(
  phiseq_df = exist,
  metadata = timepoints_df,
  ind_id_col = "ind_id",
  samples_t1_col = g1,
  samples_t2_col = g2,
  label_x = g1,
  label_y = g2,
  method = "pearson", 
  require_both_timepoints = TRUE
  )
}
```

<div style="height: 50px;"></div>


# Functional Subgroups in Antigen Library

```{r,  fig.width=12, fig.height=5}
for (group_col in group_cols) {
  comparisons <- names(comparison_group_tests[[group_col]])
  
  for (comp_name in comparisons) {
    #comp_name <- comparisons[1]
    groups <- strsplit(comp_name, "_vs_")[[1]]
    group1 <- groups[1]
    group2 <- groups[2]
    comparison_df <- comparison_group_tests[[group_col]][[comp_name]]$comparison_df[
      c("Peptide", "Organism_complete_name", group1, group2,
        paste0(group1, "_count"), paste0(group2, "_count"), "ratio")
    ]
    
    output_table <- paste0(out_dir,"/table_peptidesSignificance_",group_col,"_",comp_name,".csv")
    readr::write_csv(comparison_group_tests[[group_col]][[comp_name]]$comparison_df, output_table)
    
    p <- plot_ratios_by_subgroup(comparison_df, group1, group2, subgroup_lib_df, 
                                              custom_colors, prevalence_threshold = 0)
    
    if (!is.null(p)) {
      print(p)
      # ggsave(paste0(group_col, "_", comp_name, "_subgroup_ratios.png"), p)
    }
  }
}
```
