---
title: "Descriptive Report"
subtitle: "`r params$comparison`"
author: "Carlos S. Reyna-Blanco"
license: "CC-BY 4.0"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: true
    theme: united
    highlight: tango
    toc: true           # Enable table of contents
    toc_float: true     # Make the table of contents float on the side
    number_sections: true  # Optional: Number your sections
    toc_depth: 4  
params:
  metadata:      NULL    # will be a data.frame of samples × columns
  exist:         NULL    # will be a data.frame of peptides × samples
  comparison:    NULL    # the name of this comparison
  library_meta:  NULL    # will be a data.frame of the library metadata
  prevalence_threshold: 0 # will be filter applied on peptides based on prevalence
  custom_colors: NULL    # Colors to use for each group
  timepoints:    NULL    # Only passed if data is longitudinal and a ind_timepoint.csv is given
  out_tables:    NULL    # where to save the tables
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo  = FALSE, warning = FALSE, message = FALSE
                      ,fig.width  = 5, fig.height = 4) # default figure dimensions (in inches)

# Load custom utility functions
source(file.path("utils", "report_utils.R"))

# read in the two per‐comparison files
metadata  <- params$metadata
exist <- params$exist
comparison <- params$comparison
lib_metadata_df <- params$library_meta
prevalence_threshold <- params$prevalence_threshold
custom_colors <- params$custom_colors
out_dir <- params$out_tables



lib_metadata_df <- lib_metadata_df %>%
  mutate(
    species = str_remove_all(species, "\\[|\\]"), #a few species with this
    species = str_squish(species)     
  )

subgroup_lib_df <- lib_metadata_df %>%
  tibble::rownames_to_column(var = "Peptide") %>%
  mutate(
    across(
      all_of(SUBGROUPS_TO_INCLUDE),
      ~ if_else(is.na(.x), FALSE, .x)
    )
  ) %>%
  select(Peptide, all_of(SUBGROUPS_TO_INCLUDE))
```


```{r}

#metadata      = meta_i
#shannon_diversity <- vegan::diversity(t(exist), index = "shannon", base = 2)
#samples_stats <- data.frame(SampleName = names(shannon_diversity), enriched_peptides = colSums(exist), Shannon_diversity = shannon_diversity)
#samples_stats <- data.frame(SampleName = colnames(exist), enriched_peptides = colSums(exist))
#metadata <- metadata %>% 
#  merge(samples_stats, by = "SampleName") #, all.x = TRUE)


# find Sex and Age column names, case‑insensitive
lc <- tolower(names(metadata))
# find age and sex/gender columns
age_idx <- which(lc == "age")
sex_idx <- which(lc %in% c("sex", "gender"))

if (length(age_idx) == 1 && length(sex_idx) == 1) {
  age_col <- names(metadata)[age_idx]
  sex_col <- names(metadata)[sex_idx]
  
  metadata <- metadata %>%
    mutate(
      Age_group = cut(
        .data[[age_col]],
        breaks = seq(0, 120, by = 5),
        right  = FALSE
      ),
      
      Sex_ctg = {
        # coerce to character and lowercase once
        sex_raw <- as.character(.data[[sex_col]])
        sex_lc  <- str_to_lower(sex_raw)
        
        # map to “Female” or “Male”
        sex_norm <- case_when(
          sex_lc %in% c("female", "f", "0") ~ "Female",
          sex_lc %in% c("male",   "m", "1") ~ "Male",
          TRUE                              ~ NA_character_
        )
        
        # factor with the desired order
        factor(sex_norm, levels = c("Female", "Male"))
      }
    )
  
} else {
  warning("Could not find exactly one Age and one Sex/Gender column; skipping Sex/Age grouping.")
}

# set group and merge features with group test labels
group_cols <- c("group_test")
features_target <- as.data.frame(t(exist)) %>%
  tibble::rownames_to_column("SampleName")%>%
  dplyr::left_join(metadata %>% select(any_of(c("SampleName", group_cols))), by = "SampleName")
```


```{r, results='asis'}
required <- c("Age_group", "Sex_ctg")
show_sexAge <- all(required %in% colnames(metadata))
```


`r if (show_sexAge) "\n# Sex/Age Distribution"` 

Distribution of Sex/Age per group comparisons.

```{r, fig.height=5, fig.width=8}
if (show_sexAge) {
  sex_colors <- c("Female" = "#66C2A5", "Male" = "#FC8D62")
  p <- plot_sex_age_distribution(data = metadata, 
                                 group_col = group_cols, 
                                 age_col = "Age_group", 
                                 sex_col = "Sex_ctg", 
                                 custom_colors = sex_colors)
  print(p)
} 
```


# Antibody Epitope Repertoire of Samples


Antibody epitope repertoires of all individuals or per group recognize private (occurring in single individuals) and public (shared in all of the cohort) antigens.

- Plots shows how often a number of peptides is present only in certain number of individuals per group comparisons.

- Highlighting 5% prevalence filter and the amount of peptides that remain after the filter.

```{r, fig.height=5, fig.width=7}
for (group_col in group_cols){
  p <- plot_enrichment_counts(
    features_target      = features_target,
    group_col            = group_col,
    group_cols           = group_cols,
    prevalence_threshold = 0,
    custom_colors        = custom_colors,
    binwidth             = 0.5
  )
  print(p)
}
```


```{r}
counts_list <- generate_taxa_count_tables(lib_metadata_df, exist)
```


# Alpha Diversity  {.tabset}

-Richness: the absolute number of antibody-bound peptides/taxa rank per individual between groups.
-Shannon: A given taxonomic rank diversity of Ig repertoires of antibody-bound peptides per individual between groups.

```{r alpha‐all‐ranks, results='asis', fig.width=8, fig.height=4}
# define the ranks you want to iterate over
rank_levels <- rev(names(counts_list))[1:5]  # e.g. c("species","genus",…)

for(rank_level in c("peptide", rank_levels)) {

  cat("\n\n## ", tools::toTitleCase(rank_level), " {.tabset} \n\n")
  
  if (rank_level %in% "peptide"){
    counts <- exist
    label_y <- paste("Counts of Enriched Peptides")

  } else{
    counts <- counts_list[[rank_level]]
    label_y <- paste("Richness (", rank_level, " rank)", sep = "")
  }
  
  for (group_col in group_cols){
  
    sample_meta  <- metadata %>% select(SampleName=SampleName, all_of(group_col))
    alpha_df     <- compute_alpha_diversity(counts, sample_meta, group_col)
    pairwise_comparisons <- combn(levels(factor(sample_meta[[group_col]])), 2, simplify = FALSE)
    p1 <- plot_groups_boxplots(data = alpha_df, 
                              group_col = "Group", 
                              values_col = "Richness",
                              custom_colors = custom_colors, 
                              pairwise_comparisons = pairwise_comparisons,
                              label_axis = c("Group", label_y), 
                              sig_level = 1,label_format = "p.signif") #"p.format")
    p2 <- plot_groups_boxplots(data = alpha_df, 
                              group_col = "Group", 
                              values_col = "Shannon Diversity",
                              custom_colors = custom_colors, 
                              pairwise_comparisons = pairwise_comparisons,
                              label_axis = c("Group", paste("Shannon Diversity (", tools::toTitleCase(rank_level), " rank)",sep="")), 
                              sig_level = 1,label_format = "p.signif") #"p.format")
    # p3 <- plot_groups_boxplots(data = alpha_df, 
    #                           group_col = "Group", 
    #                           values_col = "Simpson Diversity",
    #                           custom_colors = custom_colors, 
    #                           pairwise_comparisons = pairwise_comparisons,
    #                           label_axis = c("Group Test", "Simpson's Diversity Index"), sig_level = 0.05,label_format = "p.signif")
    print(p1+p2)
  }
}
```


# Beta Diversity {.tabset}

- For each taxonomic rank (e.g. species, genus, …) Bray–Curtis dissimilarity is computed on the relative abundance taxonomic matrix and carried out a Principal Coordinates Analysis (PCoA/MDS). 

- On the scatterplot, each point is a sample and the 95 % confidence ellipses show the group clustering. Below it, the boxplot displays each sample’s distance to its group centroid.

- Tested for overall differences in community composition between groups with PERMANOVA, reporting the p-value in the plot. To check that any PERMANOVA result wasn’t driven by unequal dispersion, a permutation test on the multivariate dispersions (betadisper) was run.

- Together, these panels tell you whether (a) the two groups occupy distinct regions of Bray–Curtis space (significant PERMANOVA) and (b) whether one group is more dispersed than the other (significant dispersion test).

```{r beta‐all‐ranks, results='asis', fig.width=8, fig.height=4}
# rank <- "species"
# counts <- counts_list[[rank]] %>% 
#   tibble::rownames_to_column("Species") %>%
#   filter(Species != "Homo sapiens") %>%
#   tibble::column_to_rownames("Species")
#group_col <- "group_test"

#for(rank_level in c("peptide",rank_levels)) {
for(rank_level in c(rank_levels)) {

  cat("\n\n## ", tools::toTitleCase(rank_level), " {.tabset} \n\n")
  
  if (rank_level %in% "peptide"){
    counts <- exist
  } else{
    counts <- counts_list[[rank_level]]
  }
  for (group_col in group_cols) {
    sample_meta  <- metadata %>% 
      select(SampleName=SampleName, all_of(group_col)) #%>% rename(Group = !!sym(group_col))  
    
    if (rank_level %in% "peptide"){
      beta_res <- compute_beta_diversity(counts, sample_meta, group_col = group_col, method = "bray", norm_method = "none")
    } else{
      beta_res <- compute_beta_diversity(counts, sample_meta, group_col = group_col, method = "bray", norm_method = "relative")
    }
    
    p1 <- plot_pcoa(beta_diversity= beta_res, custom_colors = custom_colors, ellipse = T, permutations = 1000, show_legend = F )
    p2 <- plot_beta_dispersion(beta_res$beta_dispersion, custom_colors, sig_level   = 1, label_format  = "p.signif")
    print(p1+p2)
  }
}
#ggsave("my_plot.svg", plot = p1+p2, width = 6, height = 4, device = "svg")
```



# Peptide Comparison Between Groups {.tabset}

```{r}
num_samples_per_group <- list()
for (group_col in group_cols) {
  num_samples_per_group[[group_col]] <- metadata %>%
    filter(!is.na(!!sym(group_col))) %>%  # Filter out NA values in the group_test column
    group_by(!!sym(group_col)) %>%
    summarise(n = n(), .groups = "drop") %>%
    tibble::deframe()  # Convert the summarized result to a named vector (group -> count)
}

percentage_group_test_list <- list()
for (group_col in group_cols){
  percentage_group_test_list[[group_col]] <-  get_count_percentage_df(
    features_target      = features_target,
    group_col            = group_col,
    group_cols           = group_cols,
    lib_metadata_df      = lib_metadata_df,
    prevalence_threshold = 0)
}

prevalence_threshold <- prevalence_threshold
comparison_group_tests <- group_test_analysis(percentage_group_test_list, num_samples_per_group, prevalence_threshold)
```


## Significance (Peptide level)   {.tabset}


- Scatterplots show in green peptides that are significant before multiple testing correction and in red peptides that are significant after correction (BH)

- pvals were calculated using Fisher test comparing peptide prevalence between groups


### Scatterplot {.tabset}


```{r sig-scatter-peptide, results='asis'}
#group <- names(comparison_group_tests)[1]
all_plots <- lapply(comparison_group_tests[[group_col]], function(x) {
  # Adjusting the plot margins to avoid hover text cropping
  x$plot <- htmltools::div(style = "margin-bottom: 150px;", x$plot)
})
htmltools::tagList(all_plots)
```
<div style="height: 50px;"></div>



### Volcano plot {.tabset}

- A different way to visualize  significant peptides based on the prevalence ratio of two groups and corresponding pvalues (before multiple test correction)
- pvals were calculated using Fisher test comparing peptide prevalence between groups

```{r sig-volcano-peptide, results='asis'}
volcano_plots <- list()
for (group_col in group_cols) {
  comparisons <- names(comparison_group_tests[[group_col]])
  for (comp_name in comparisons) {
    #comp_name <- comparisons[1]
    groups <- strsplit(comp_name, "_vs_")[[1]]
    group1 <- groups[1]
    group2 <- groups[2]
    comparison_df <- comparison_group_tests[[group_col]][[comp_name]]$comparison_df
    p <- make_interactive_volcano(comparison_df = comparison_df, group1 = group1, group2 = group2)
    volcano_plots[[paste0(group1, "_vs_", group2)]] <- list(plot = p %>% htmltools::div(style = "margin-bottom: 150px;"))
    # ggsave(paste0(group_col, "_", comp_name, "_volcano_plot.png"), p)
  }
  print(htmltools::tagList(volcano_plots))
}
```


## Significance (Taxa level)  {.tabset}

- Highlighting interesting virus/bacteria groups at rank level that seems to appear significantly overrepresented when compared to the complete library

- Adjusted pvals (BH) were calculated using MW test comparing the ratio of group1/group2 of the bacteria/virus rank of interest and the ratio of group1/group2 of the complete library (excluding the group of interest)

```{r}
#highlight_colors <- c("#7570b3", "cadetblue3", "gold", "lightsalmon2", "chartreuse4", "magenta4", "royalblue2", "tomato2", "yellow4", "springgreen2")
#highlight_colors <- c("#7570b3", "royalblue2", "gold", "tomato2", "chartreuse4", "magenta4", "cyan3", "pink3", "yellowgreen", "sienna4")
highlight_colors <- c("#7570b3", "#1b9e77", "gold", "tomato2", "royalblue2", "chartreuse4", "magenta4", "cyan3", "pink3", "sienna4")
significant_taxa_list <- list()
```

### Scatterplot  {.tabset}

```{r sig-scatter‐all‐ranks, results='asis'}
results_list <- list()
for(rank_level in rank_levels) {
  #rank_level <- "species"
  cat("\n\n#### ", tools::toTitleCase(rank_level), " {.tabset} \n\n")
  significant_taxa_list[[rank_level]] <- list()
  for (group_col in group_cols) {
    scatter_plots <- list()
    comparisons <- names(comparison_group_tests[[group_col]])
    for (comp_name in comparisons) {
      #comp_name <- comparisons[1]
      groups        <- strsplit(comp_name, "_vs_")[[1]]
      group1        <- groups[1]
      group2        <- groups[2]
      N1 <- num_samples_per_group[[group_col]][[group1]]
      N2 <- num_samples_per_group[[group_col]][[group2]]
      comparison_df <- comparison_group_tests[[group_col]][[comp_name]]$comparison_df
     
      significant_taxa_list[[rank_level]][[comp_name]] <- make_flag_lists_from_top_names(comparison_df = comparison_df, lineage_col = rank_level, 
                                                                                         n = 10, colors = highlight_colors)


      all_flags <- names(significant_taxa_list[[rank_level]][[comp_name]]$flags_to_patterns)
      # if there are no flags at all, skip this comparison
      if (length(all_flags) == 0) {
        message("  → no significant taxa for ", comp_name, "; skipping.")
        next
      }
  
      # otherwise take up to five
      flags    <- head(all_flags, 5)
      n_flags  <- length(flags)
      for(flag in flags){
        patterns <- significant_taxa_list[[rank_level]][[comp_name]]$flags_to_patterns[[flag]]
        # (A) Add a new TRUE/FALSE column "flag" by matching those patterns:
        comparison_df <- comparison_df %>%
          add_exact_flag(
            new_flag    = flag,
            patterns    = patterns,
            target_cols = c(rank_level, "Description")
          )
      }
      p <- make_interactive_scatterplot(comparison_df = comparison_df,
                               group1 = group1, group2 = group2, N = c(N1,N2),
                               highlight_cols   = flags, 
                               #highlight_colors = c(Bacteriodales = "#1b9e77", Enterovirus = "#7570b3"),
                               highlight_colors = significant_taxa_list[[rank_level]][[comp_name]]$highlight_colors[1:n_flags],
                               default_color    = "gray80", 
                               interactive = T)
      scatter_plots[[paste0(group1, "_vs_", group2,"-",flag)]] <- list(plot = p)
    }
    results_list[[group_col]] <- scatter_plots
  }

  all_plots <- lapply(results_list[[group_cols[1]]], function(x) {
    # Adjusting the plot margins to avoid hover text cropping
    x$plot <- htmltools::div(style = "margin-bottom: 150px;", x$plot)
  })
  print(htmltools::tagList(all_plots))
}
```


### Volcano plot {.tabset}

```{r sig-volcano‐all‐ranks, results='asis'}
results_list <- list()
for(rank_level in rank_levels) {
  cat("\n\n#### ", tools::toTitleCase(rank_level), " {.tabset} \n\n")
  for (group_col in group_cols) {
    volcano_plots <- list()
    comparisons <- names(comparison_group_tests[[group_col]])
    for (comp_name in comparisons) {
      #comp_name <- comparisons[1]
      groups        <- strsplit(comp_name, "_vs_")[[1]]
      group1        <- groups[1]
      group2        <- groups[2]
      N1 <- num_samples_per_group[[group_col]][[group1]]
      N2 <- num_samples_per_group[[group_col]][[group2]]
      comparison_df <- comparison_group_tests[[group_col]][[comp_name]]$comparison_df

      all_flags <- names(significant_taxa_list[[rank_level]][[comp_name]]$flags_to_patterns)
      # if there are no flags at all, skip this comparison
      if (length(all_flags) == 0) {
        message("  → no significant taxa for ", comp_name, "; skipping.")
        next
      }
  
      # otherwise take up to five
      flags    <- head(all_flags, 10)
      n_flags  <- length(flags)
      for(flag in flags){
        patterns <- significant_taxa_list[[rank_level]][[comp_name]]$flags_to_patterns[[flag]]
        # (A) Add a new TRUE/FALSE column "flag" by matching those patterns:
        comparison_df <- comparison_df %>%
          add_exact_flag(
            new_flag    = flag,
            patterns    = patterns,
            target_cols = c(rank_level, "Description")
          )
      }
      # flag_counts <- sapply(flags, function(f) {
      #   n_true  <- sum(comparison_df[[f]], na.rm = TRUE)
      #   n_false <- sum(!comparison_df[[f]], na.rm = TRUE)
      #   c(n_true = n_true, n_false = n_false)
      #   })
      # valid_flags <- names(which(
      #   flag_counts["n_true", ]  >= 2 &
      #   flag_counts["n_false", ] >= 2
      # ))
      # if (length(valid_flags) == 0) {
      #   message("No flags with ≥2 cases and ≥2 non‐cases—skipping this comparison.")
      #   next
      # }
      # valid_colors <- significant_taxa_list[[rank_level]][[comp_name]]$highlight_colors[valid_flags]
      p <- make_interactive_volcano(comparison_df = comparison_df,
                               group1 = group1, group2 = group2, fc_cut = 1, p_cut = 0.05,
                               highlight_cols   = flags,  
                               highlight_colors = significant_taxa_list[[rank_level]][[comp_name]]$highlight_colors[1:n_flags],
                               default_color    = "gray80", 
                               interactive = T)
      volcano_plots[[paste0(group1, "_vs_", group2,"-",flag)]] <- list(plot = p)
    }
    results_list[[group_col]] <- volcano_plots
  }
  
  all_plots <- lapply(results_list[[group_cols[1]]], function(x) {
    # Adjusting the plot margins to avoid hover text cropping
    x$plot <- htmltools::div(style = "margin-bottom: 20px;", x$plot)
  })
  print(htmltools::tagList(all_plots))
}
```


## Table

```{r, results='asis'}
DT::datatable(
  head(comparison_group_tests[[group_cols[1]]][[comparison]]$comparison_df, 300),
  options = list(
    pageLength = 5,
    scrollX = TRUE,  # enable horizontal scrolling
    width = "100%"
  ),
  caption = "Preview of the top 300 Peptides ordered by delta ratio and pvals"
)
```
<div style="height: 50px;"></div>



```{r, results='asis'}
show_timepoints <- !is.null(params$timepoints)
if (show_timepoints) {
  cat('\n\n<div style="height: 60px;"></div>\n\n')  # adds vertical space
  cat('\n\n<div style="height: 60px;"></div>\n\n')  # adds vertical space
}

```


`r if (show_timepoints) "\n\n# Longitudinal Similarity {.tabset}"`  
`r if (show_timepoints) "## Data in at least one timepoint"`  


```{r corr-plot, fig.height=9, fig.width=9}
if (show_timepoints) {
timepoints_df <- params$timepoints
g1 <- levels(factor(metadata[[group_col]]))[1]
g2 <- levels(factor(metadata[[group_col]]))[2]
  
plot_correlation(
  phiseq_df = exist,           # assuming your function uses this as the data matrix
  metadata = timepoints_df,
  ind_id_col = "ind_id",
  samples_t1_col = g1,
  samples_t2_col = g2,
  label_x = g1,
  label_y = g2,
  method = "pearson", 
  require_both_timepoints = FALSE
  )
}
```


`r if (show_timepoints) "## Data in both timepoints"`  


```{r both-corr-plot, fig.height=9, fig.width=9}
if (show_timepoints){
timepoints_df <- params$timepoints
g1 <- levels(factor(metadata[[group_col]]))[1]
g2 <- levels(factor(metadata[[group_col]]))[2]
  
plot_correlation(
  phiseq_df = exist,           # assuming your function uses this as the data matrix
  metadata = timepoints_df,
  ind_id_col = "ind_id",
  samples_t1_col = g1,
  samples_t2_col = g2,
  label_x = g1,
  label_y = g2,
  method = "pearson", 
  require_both_timepoints = TRUE
  )
}
```


`r if (show_timepoints) "\n\n#  Random and Matched Pairs of Samples Similarity Distribution {.tabset}"`  
`r if (show_timepoints) "## Data in at least one timepoint"`


```{r corr-plot-dist, fig.height=6, fig.width=6}
if (show_timepoints){
timepoints_df <- params$timepoints
g1 <- levels(factor(metadata[[group_col]]))[1]
g2 <- levels(factor(metadata[[group_col]]))[2]

    
plot_correlation_distribution(
  phiseq_df = exist,           # assuming your function uses this as the data matrix
  metadata = timepoints_df,
  ind_id_col = "ind_id",
  samples_t1_col = g1,
  samples_t2_col = g2,
  label_x = g1,
  label_y = g2,
  method = "pearson", 
  require_both_timepoints = FALSE
  )
}
```


`r if (show_timepoints) "## Data in both timepoints"`  


```{r both-corr-plot-dist, fig.height=6, fig.width=6}
if (show_timepoints){
timepoints_df <- params$timepoints
g1 <- levels(factor(metadata[[group_col]]))[1]
g2 <- levels(factor(metadata[[group_col]]))[2]
 
plot_correlation_distribution(
  phiseq_df = exist,
  metadata = timepoints_df,
  ind_id_col = "ind_id",
  samples_t1_col = g1,
  samples_t2_col = g2,
  label_x = g1,
  label_y = g2,
  method = "pearson", 
  require_both_timepoints = TRUE
  )
}
```

<div style="height: 50px;"></div>


# Subgroups in Antigen Library {.tabset}

- Represent the main antigen subgroups ratio distribution within the library (including bacteria/virus of interest) that exhibits differential binding between groups.

- If ratio is positive between group1/group2, this indicate that group1 (eg. Controls) shows more representation of that peptide subgroup than group2 (e.g HCC)

- **This is a stat testing approach, might not work well for small sample sizes or be biased**

- **By default, a 0% prevalence peptide filter is applied (user can defined it) (only dropping peptides below that threshold in both groups)**


```{r sig-subgroups‐all‐ranks, results='asis', fig.width=8, fig.height=8}
for (rank_level in rank_levels){
  cat("\n\n## ", tools::toTitleCase(rank_level), " {.tabset} \n\n")
  for (group_col in group_cols[1]) {
    comparisons <- names(comparison_group_tests[[group_col]])
    for (comp_name in comparisons) {
      #comp_name <- comparisons[1]
      groups <- strsplit(comp_name, "_vs_")[[1]]
      group1 <- groups[1]
      group2 <- groups[2]
      comparison_df <- comparison_group_tests[[group_col]][[comp_name]]$comparison_df[
        c("Peptide", rank_level, "Description", group1, group2,
          paste0(group1, "_count"), paste0(group2, "_count"), "ratio")
      ]
      
      
      all_flags <- names(significant_taxa_list[[rank_level]][[comp_name]]$flags_to_patterns)
      # if there are no flags at all, skip this comparison
      if (length(all_flags) == 0) {
        message("  → no significant taxa for ", comp_name, "; skipping.")
        next
      }
      
      flags <- names(significant_taxa_list[[rank_level]][[comp_name]]$flags_to_patterns)
      for(flag in flags){
        patterns <- significant_taxa_list[[rank_level]][[comp_name]]$flags_to_patterns[[flag]]
        col_flag <- comparison_df %>%
          add_exact_flag(
          new_flag    = flag,
          patterns    = patterns,
          target_cols = c(rank_level, "Description")
        )  %>% 
          select(Peptide, all_of(flag)) %>%
          distinct()
        
        # Add a new TRUE/FALSE column "flag" by matching those patterns:
        subgroup_lib_df <- subgroup_lib_df %>% 
          select(-any_of(flag)) %>%                # remove the column named by `flag` if it exists
          left_join(col_flag, by = "Peptide")
      }
      
      if (rank_level == "species"){
        output_table <- paste0(out_dir,"/table_peptidesSignificance_",group_col,"_",comp_name,".csv")
        readr::write_csv(comparison_group_tests[[group_col]][[comp_name]]$comparison_df, output_table)
      }
      
      p_subgroup <- plot_ratios_by_subgroup(comparison_df, group1, group2, subgroup_lib_df, 
                                   prevalence_threshold = prevalence_threshold,
                                   which_subgroups = "default"
                                   )
      p_taxa <- plot_ratios_by_subgroup(comparison_df, group1, group2, subgroup_lib_df, 
                                   subgroup_colors = significant_taxa_list[[rank_level]][[comp_name]]$highlight_colors,
                                   add_subgroups = flags,
                                   prevalence_threshold = prevalence_threshold,
                                   which_subgroups = "added",
                                   x_label = paste("Significant", rank_level, "of antigen library", sep = " ")
                                   )
      # venn_p <- plot_ratio_venn(comparison_df, 
      #                 group1,
      #                 group2,
      #                 custom_colors)
      combined_grid <- (p_subgroup$boxplot + p_subgroup$heatmap + plot_layout(widths = c(1,1))) /
        (p_taxa$boxplot     + p_taxa$heatmap     + plot_layout(widths = c(1,1))) +
        plot_layout(guides = "collect") &
        theme(legend.position = "bottom")
  
      print(combined_grid)
  
    }
  }
}
```




# Significant Taxa Between Groups {.tabset}


```{r significant‐all‐ranks, results='asis', fig.width=8, fig.height=8}
for(rank_level in rank_levels) {

  cat("\n\n## ", tools::toTitleCase(rank_level), " {.tabset} \n\n")
  
  preps <- lapply(group_cols, function(gc) {
    prep <- prep_taxa_stats(
      rank        = rank_level,
      group_col   = gc,
      counts_list = counts_list,
      metadata    = metadata,
      top_n       = 12,
      order_by    = "All"    # or specify e.g. "Controls"
    )
  })
  names(preps) <- group_cols


  for(group_col in group_cols) {
    
    prep <- preps[[group_col]]
      
    # get the max abundance
    max_ab <- max(prep$rel_abundance_filtered$Abundance, na.rm = TRUE)
    
    # choose your nudge based on thresholds
    nudge_y <- case_when(
      max_ab <= 0.1   ~ -1.0,
      max_ab <= 0.15  ~ -0.8,
      max_ab <= 0.2   ~ -0.7,
      max_ab <= 0.3   ~ -0.5,
      TRUE           ~ -0.3    # a sensible default if >0.3
    )

    if (nrow(prep$pval_df) > 0) {
      # Proceed with plotting
      p1 <- plot_taxa_violin(prep, ncol = 3, nudge_y = nudge_y)
      print(p1)
    } else {
      message("No data to plot.")
    }
    

    # bar_df <- prep$rel_abundance_filtered %>%
    #   group_by(Rank, Group) %>%
    #   summarise(
    #     median_ab = median(Abundance[Abundance > 0], na.rm = TRUE),
    #     #mean_ab = median(Abundance, na.rm = TRUE),
    #     #sd_ab   = sd(Abundance,   na.rm = TRUE),
    #     .groups = "drop"
    #     ) %>%
    #   mutate(median_ab = median_ab * 100)
    #   #sd_ab   = sd_ab   * 100)
    # 
    # 
    # p2 <- ggplot(bar_df, aes(x = Rank, y = median_ab, fill = Group)) +
    #   geom_col(position = position_dodge(width = 0.8), width = 0.7) +
    # 
    #   scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
    #   scale_fill_manual(values=custom_colors) +
    #   labs(
    #     x = tools::toTitleCase(rank_level),
    #     y = "Median Relative Abundance (%)",
    #     fill = "Group"
    #   ) +
    #   theme_pubclean(base_size = 12) +
    #   theme(
    #     axis.text.x = element_text(angle = 0, hjust = 0.5)
    #   ) +
    #   coord_flip()
    # 
    # print(p2)
  }
}

```

